Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-09-15 ms/ms.qmd
reformat ms

---
title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gias}{$g_\text{ias}$}
  \newcommand{\gmax}{$g_\text{max}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: science.csl
---

```{=latex}
\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}
```

@@

@@

@@

>>>>

---
title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gias}{$g_\text{ias}$}
  \newcommand{\gmax}{$g_\text{max}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
  \usepackage{setspace}
  \usepackage[modulo]{lineno}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: https://www.zotero.org/styles/pnas
---

```{=latex}
\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\medskip
Classification: Biological Sciences, Ecology\\
\medskip
Keywords: amphistomy, photosynthesis, \textit{Solanum}, stomata
\end{center}
```

@@

```{=latex}
\newpage
\begin{linenumbers}
\begin{doublespace}
```

@@

## Significance statement

Stomata, the "mouths" of plants, regulate CO$_2$ available for photosynthesis. They are ubiquitous on leaves and essential to understanding why vascular plants transformed Earth and how our crops grow. Yet the adaptive significance of why many leaves only have stomata on their lower surface is a mystery. Many assume that this limits water loss, but we show the opposite is true. By manipulating which stomata a leaf can use, we show that leaves with stomata on both surfaces gain more CO$_2$ for the same amount of water lost, increasing water use efficiency. Furthermore, species from open canopies grown under high light benefit the most, which may explain why leaves with stomata on both sides are most prevalent in these habitats.

@@

```{=latex}
\end{doublespace}
\end{linenumbers}
```
